---
title: "Using `rJavaEnv` in R Packages"
vignette: >
  %\VignetteIndexEntry{Using `rJavaEnv` in R Packages}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  eval: false
format:
  html:
    toc: true
    toc-depth: 2
    code-overflow: wrap
---

If you are developing an R package that involves `Java` (either via `{rJava}` or through command-line `Java` tools), you often face the "Java Hell" problem: your users might not have Java installed, or they might have the wrong version, or `{rJava}` cannot find the `JAVA_HOME`.

`rJavaEnv` solves this by providing a unified, reliable way to help your users ensure a specific Java version is present and configured correctly.

This guide explains how to integrate `rJavaEnv` into your package to create a seamless experience for your users while adhering to CRAN policies and best practices.

# Critical: Understanding rJava's Path-Locking Behavior

Before deciding how to use `rJavaEnv` in your package, you must understand a fundamental limitation of the `{rJava}` package:

**Once `{rJava}` is loaded or initialized in an R session, the Java version is locked. It cannot be changed without restarting the R session.**

This is documented in `{rJava}` GitHub issues [#25](https://github.com/s-u/rJava/issues/25), [#249](https://github.com/s-u/rJava/issues/249), and [#334](https://github.com/s-u/rJava/issues/334). What this means:

- You **cannot** set Java in your package's `.onLoad` hook
  - Setting Java in `.onLoad` would lock your users to that specific Java path every time they load your package, and they won't be able to change it without restarting R and injectiong a different Java path manually into shell environment that starts R, or without setting these paths in project-level `.Renviron` or `.Rprofile` files
  - It defeats the purpose of `rJavaEnv`'s flexibility
- You **must** instruct users to set `Java` paths **before** loading your package (or any package that imports `{rJava}`)
- If a user loads your package without configuring `Java` first, they must restart R and configure Java before trying again

This is why the typical ".onLoad check" pattern used by other R packages does **not** work well for packages that depend on `{rJava}`.

### When Does rJava Get Initialized?

`{rJava}` gets locked to its Java version when any of these occur:
1. User explicitly calls `library(rJava)`
2. User loads any package that imports `{rJava}` (which auto-loads it as a dependency)
3. User's IDE autocompletes with `rJava::` (this triggers initialization!)
4. Any `{rJava}`-dependent function is called

Once any of these happen, the Java version is locked for that session.

# Two Development Scenarios

How you use `rJavaEnv` depends on whether your package uses `{rJava}` or not.

## Scenario A: Packages Using Command-Line Java (No rJava)

If your package calls `Java` via system commands (e.g., `system2("java", ...)` or perhaps via `{processx}`) and does **not** import `{rJava}`, you have flexibility.

### Strategy: "Ensure and Call"

You can use `java_ensure()` within your package functions to set the `Java` environment before each system call:

```{r}
#' Run my Java tool
#' @export
run_my_java_tool <- function(java_version = 21) {
  # Ensure the correct Java version is set before calling the tool
  rJavaEnv::java_ensure(version = java_version, type = "min", install = FALSE)

  # Now call the Java tool via system command
  system2("java", c("-jar", "my-tool.jar", "--input", "data.txt"))
}
```

**Performance Note**: `rJavaEnv` automatically caches version checks, so calling `java_ensure()` repeatedly has minimal overhead (<1ms per call after the first check). You do not need to implement your own caching logic.

This approach:
- Checks if the correct Java version is available
- Does **not** install automatically (respects `install = FALSE` to avoid network calls in functions)
- Returns `FALSE` if the required Java is not available, allowing your function to handle it gracefully
- Works seamlessly because command-line Java is not locked like `{rJava}` is

#### Understanding Version Matching: `type = "exact"` vs `type = "min"`

The `type` parameter controls how strictly Java versions must match:

**`type = "exact"` (default):** Requires an exact version match
```{r}
java_ensure(version = 21, type = "exact")  # Only Java 21 will satisfy this
```
Use this when your tool requires a specific Java version and won't work reliably with newer versions.

**`type = "min"`:** Accepts the specified version or any newer version
```{r}
java_ensure(version = 21, type = "min")  # Java 21, 22, 23, etc. will all work
```
Use this when your tool works with Java 21 and is compatible with newer versions.

#### Controlling Whether to Use System Java: `accept_system_java`

By default, `accept_system_java = TRUE`, which means:
- `rJavaEnv` will first check if a suitable Java version already exists on the user's system
- If a matching system Java is found, **it will be used immediately—no download occurs**
- If no suitable system Java is found, only then will `rJavaEnv` download and cache a version (if `install = TRUE`)

This is efficient because most users already have Java installed, so in the common case there's no network download:

```{r}
# Prefer system Java, only download if needed
java_ensure(version = 21, type = "min", accept_system_java = TRUE, install = TRUE)
```

If you want `rJavaEnv` to ignore system Java and only use versions it manages (e.g., for reproducibility or to avoid version confusion), set `accept_system_java = FALSE`:

```{r}
# Ignore system Java; must have rJavaEnv-managed Java or download one
java_ensure(version = 21, type = "min", accept_system_java = FALSE, install = TRUE)
```

### Adding rJavaEnv to DESCRIPTION

Add `rJavaEnv` to your `Imports` field:

```dcf
Imports:
    rJavaEnv
```

Your users will install `rJavaEnv` automatically when they install your package.



## Scenario B: Packages Importing rJava (Most Common)

If your package imports `{rJava}` (it's in your `Imports` field in `DESCRIPTION`), you **cannot** set Java in your package's `.onLoad` hook.

### Strategy: "Inform and Guide"

Instead, you must provide clear instructions for your users to configure Java **before** loading your package.

#### Configuration Options for Your Users

When documenting setup steps for your users, you can offer flexibility by explaining the `type` and `accept_system_java` parameters:

**Option 1: Minimum version requirement (recommended)**
```r
java_ensure(version = 21, type = "min")
```
This accepts Java 21 or newer, and reuses system Java if available. Most flexible for users.

**Option 2: Exact version requirement**
```r
java_ensure(version = 21, type = "exact")
```
Use only if your package truly requires a specific version. Stricter but potentially problematic if the user has a newer version.

**Option 3: Trust system Java, fall back to rJavaEnv-managed (default)**
```r
java_ensure(version = 21, type = "min", accept_system_java = TRUE, install = TRUE)
```
- First checks system Java (no download if found)
- Only downloads if system Java doesn't meet requirements
- Most efficient: leverages existing system installations

**Option 4: Use only rJavaEnv-managed Java (reproducibility)**
```r
java_ensure(version = 21, type = "min", accept_system_java = FALSE, install = TRUE)
```
- Ignores system Java entirely
- Ensures a predictable, rJavaEnv-controlled environment
- Useful for reproducible research or when system Java versions are unreliable

#### 1. Add rJavaEnv to Your DESCRIPTION

```dcf
Imports:
    rJava,
    rJavaEnv
```

#### 2. Document the Setup Requirement

Add a section to your README or in a vignette explaining the Java requirement. Here's a template you can customize:

```markdown
## Java Requirement

This package requires Java {YOUR_VERSION} (or higher).

**If you encounter "Java not found" or version mismatch errors**, please follow these steps:

### First-Time Setup

1. Install Java using `rJavaEnv`:

```r
install.packages("rJavaEnv")
library(rJavaEnv)
java_ensure(version = 21, type = "min")  # Replace 21 with your required version
```

2. **Restart your R session** (Session > Restart R in RStudio)

3. Then install your package normally:

```r
install.packages("your_package_name")
```

### If You Already Have Java Installed

If you already have Java installed on your system, just restart R and your package will find it automatically.

### Troubleshooting

**"Java is not installed or the wrong version is detected"**

This usually happens because:
1. Java was not configured before loading the package
2. You installed Java but didn't restart your R session

**Solution:**
```r
library(rJavaEnv)
java_ensure(version = 21, type = "min")  # Use your package's required version
# Restart R
# Try loading your package again
```

**"Still having issues?"**

Try explicitly setting the Java version:
```r
library(rJavaEnv)
use_java(21)  # Use your package's required version
# Restart R
# Try loading your package again
```

#### 3. Optional: Allow User-Controlled Java Installation

If you want to give users the option to automatically download and install Java (with explicit consent), you can implement this **outside** of `.onLoad` using `java_ensure()` with `install = TRUE`. This is CRAN-compliant as long as:

1. Users explicitly opt-in (not automatic)
2. You document how to clear the cache
3. Your package declares `Depends: R (>= 4.0)` in DESCRIPTION (rJavaEnv already does this)

Example setup function:

```{r}
#' Set up Java for this package
#' @description
#' Helper function for users to ensure Java is installed and configured
#' before using this package. Only downloads Java if the required version
#' is not already available.
#' @param java_version The Java version required (e.g., 21)
#' @export
setup_java <- function(java_version = 21) {
  if (!rJavaEnv::java_ensure(
    version = java_version,
    type = "min",
    install = FALSE
  )) {
    # Java not found, ask user
    response <- readline(
      paste0(
        "Java ", java_version, "+ not found. Download and install now? (y/n): "
      )
    )

    if (tolower(response) == "y") {
      rJavaEnv::java_ensure(version = java_version, type = "min", install = TRUE)
      message(
        "Java installed successfully! ",
        "Please restart your R session before using this package."
      )
    } else {
      stop(
        "Java ", java_version, "+ is required. ",
        "To install later, run: library(rJavaEnv); java_ensure(version = ",
        java_version, ", type = 'min')"
      )
    }
  } else {
    message("Java ", java_version, "+ is already available.")
  }
}
```

Users would then call:
```r
library(your_package)
setup_java()
# Restart R
library(your_package)  # Now ready to use
```

#### 4. Optional: Add Startup Message (if helpful)

If you'd like to add an informative message when your package loads, you can use `.onLoad` for **information only** (not to set Java):

```{r}
.onLoad <- function(libname, pkgname) {
  # Just check if Java is available (for information/diagnostics)
  java_available <- rJavaEnv::java_check_version_cmd(quiet = TRUE)

  if (isFALSE(java_available)) {
    packageStartupMessage(
      "Java 21+ was not detected in your PATH.\n",
      "To configure Java, run:\n\n",
      "  library(rJavaEnv)\n",
      "  java_ensure(version = 21, type = 'min')\n\n",
      "Then restart your R session before using this package."
    )
  }
}
```

**Important:** Do **not** call `java_ensure()` with `install = TRUE` in `.onLoad`. Automatic installation in `.onLoad` violates CRAN policies and won't help because `{rJava}` is already locked by the time `.onLoad` runs.

#### 5. Documenting Cache Management

If your package uses `java_ensure(..., install = TRUE)`, users should know how to manage the cache. Add this to your README or vignette:

> **Managing Java Cache**
>
> When this package downloads Java via `rJavaEnv`, it caches the downloaded files and installations in your system's data directory to avoid re-downloading. If you need to clear this cache to free up disk space, you can use:
>
> ```r
> library(rJavaEnv)
>
> # View installed Java versions in cache
> java_list("installed")
>
> # View downloaded distributions in cache
> java_list("distrib")
>
> # Clear all cached installations
> java_clear("installed", delete_all = TRUE)
>
> # Clear all downloaded distributions
> java_clear("distrib", delete_all = TRUE)
> ```
>
> Both commands will prompt for confirmation before deleting. Use `delete_all = TRUE` to skip the confirmation prompt.

# Updating Your DESCRIPTION

For both scenarios, add `rJavaEnv` to your `Imports` field in `DESCRIPTION`:

```dcf
Imports:
    rJavaEnv
```

You can optionally note the Java requirement in `SystemRequirements`:

```dcf
SystemRequirements: Java (>= 21)
```

# Performance & Caching

`rJavaEnv` functions use transparent session-scoped caching to improve performance, especially important for packages that call these functions frequently.

## How Caching Works

By default, caching is **disabled** (`.use_cache = FALSE`) for safety and correctness. This means:
- Every call performs a fresh check
- Safe for interactive use and one-off checks
- Slightly slower, but guaranteed to reflect current state

For performance-sensitive code (e.g., loops or initialization code), you can enable caching:

```{r}
# One-time expensive operation with caching enabled
java_check_version_cmd(.use_cache = TRUE)  # 36.7ms on first call
java_check_version_cmd(.use_cache = TRUE)  # <1ms on subsequent calls
```

## Performance Benchmarks

Here are typical timings with caching enabled (`.use_cache = TRUE`):

| Function | First Call | Cached Calls | Use Case |
|----------|-----------|--------------|----------|
| `java_check_version_cmd()` | ~37ms | <1ms | Check Java version via command-line |
| `java_check_version_rjava()` | ~301ms | <1ms | Check Java version via rJava subprocess |
| `java_find_system()` | ~209ms | <1ms | Scan system for all Java installations |

## When to Enable Caching

### Enable `.use_cache = TRUE` in:
- Package initialization code (`.onLoad` or similar)
- Loops that call the same checking function multiple times
- Performance-critical paths where version checks are called frequently

Example:
```{r}
# In your package's .onLoad hook (for diagnostics only)
.onLoad <- function(libname, pkgname) {
  # Safe use of caching in initialization - no Java locking here
  java_version <- rJavaEnv::java_check_version_cmd(quiet = TRUE, .use_cache = TRUE)

  if (isFALSE(java_version)) {
    packageStartupMessage(
      "Java was not detected. Run:\n",
      "  library(rJavaEnv); java_ensure(version = 21)\n",
      "Then restart your R session."
    )
  }
}
```

### Keep default `.use_cache = FALSE` for:
- Interactive use (one-off checks in the console)
- General-purpose function calls that aren't time-critical
- Any situation where you need current data reflecting recent Java installations

## Cache Behavior Details

**Session-scoped**: Caches are cleared when you restart R. Installing new Java during a session won't be reflected by cached results.

**Key-based for version checks**: `java_check_version_cmd()` caches per `JAVA_HOME` value. Changing `JAVA_HOME` creates a new cache entry:
```{r}
java_check_version_cmd(.use_cache = TRUE)  # Checks Java at /usr/lib/jvm/java-21/
# ... later, JAVA_HOME changes to /usr/lib/jvm/java-17/
java_check_version_cmd(.use_cache = TRUE)  # Fresh check with new JAVA_HOME
```

**System-wide for scanning**: `java_find_system()` caches all system Java installations and always recalculates the `is_default` flag to reflect the current `JAVA_HOME`.

# CI/CD (GitHub Actions)

When testing your package on GitHub Actions, ensure Java is available before your tests run.

## Option A: Use the Standard setup-java Action (Recommended)

`rJavaEnv` will detect this system Java automatically:

```yaml
steps:
  - uses: actions/checkout@v4
  - uses: actions/setup-java@v4
    with:
      distribution: 'corretto'
      java-version: '21'
  - uses: r-lib/actions/setup-r@v2
  - uses: r-lib/actions/setup-r-dependencies@v2
  - uses: r-lib/actions/check-r-package@v2
```

## Option B: Let `rJavaEnv` Handle Java Setup

If you prefer not to use `setup-java`, you can install Java via `rJavaEnv` before your tests:

```yaml
steps:
  - uses: actions/checkout@v4
  - uses: r-lib/actions/setup-r@v2
  - uses: r-lib/actions/setup-r-dependencies@v2
  - name: Install Java via rJavaEnv
    run: |
      install.packages("rJavaEnv")
      rJavaEnv::java_ensure(version = 21, type = "min")
    shell: Rscript {0}
  - uses: r-lib/actions/check-r-package@v2
```

# Summary

| Scenario | Java Source | Setup Method | User Workflow |
|----------|-------------|--------------|---------------|
| **Command-line Java (no rJava)** | System or cached | Use `java_ensure()` in functions | Install package, Java is managed per-call |
| **rJava-dependent packages** | System or cached | Users set up before loading | `java_ensure()`, restart R, then install package |

**Key Principle**: Respect rJava's path-locking constraint. Never try to set Java dynamically after rJava is loaded. Instead, guide your users to configure it upfront.

For more details on rJava's limitation, see [rJava issues #25](https://github.com/s-u/rJava/issues/25), [#249](https://github.com/s-u/rJava/issues/249), and [#334](https://github.com/s-u/rJava/issues/334).
