---
title: "Step-by-step: Download, Install, and Setup `Java` for 'R' Projects"
vignette: >
  %\VignetteIndexEntry{Step-by-step: Download, Install, and Setup `Java` for 'R' Projects}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  eval: false
format:
  html:
    toc: true
    toc-depth: 2
    code-overflow: wrap
---


The basics of using `rJavaEnv` are covered in the [Quick Start Guide](rJavaEnv.qmd), which demonstrates how to install Java in one line of code. This guide will show you how to have a more fine-grained control over your Java environments. Also see the [vignette on using `rJavaEnv` with the `targets` and `callr` packages to manage multiple `Java` environments](multiple-java-with-targets-callr.qmd).

```{r}
library(rJavaEnv)
```

Assume your project directory is currently in a temporary directory. Feel free to skip that, if you are already working in a desired project directory where you would like to install `Java`. In the example outputs below you will see paths that point to a temporary directory, but in a real project you would see your project directory instead.

```{r}
project_dir <- tempdir()
setwd(project_dir)
```


# 1. Set-up cache folder location


You can specify the cache folder location with the `rJavaEnv.cache_path` option.

```{r}
#| eval: false
options(rJavaEnv.cache_path = "/path/to/your/desired/cache/folder")
```

```{r}
#| include: false
options(rJavaEnv.cache_path = tempdir()) # does that fix the error on r-oldrel-macos-arm64 and r-oldrel-macos-x86_64?
```

By default the location would be set to whichever path is returned by `tools::R_user_dir("rJavaEnv", which = "cache")`. The exact path will depend on the operating system. Our recommendation is to leave this option at it's default setting. This is similar to how `renv` package manages the cache for `R` packages.

You can view the current cache location with:

```{r}
getOption("rJavaEnv.cache_path")
```

Expected output (on a Windows machine):

```
"C:\\Users\\user_name\\AppData\\Local/R/cache/R/rJavaEnv"
```

Expected output (on a macOS machine):

```
"/Users/user_name/Library/Caches/org.R-project.R/R/rJavaEnv"
```

# 2. List available Java versions

Before downloading, you can check which versions are available. By default, `rJavaEnv` checks specific vendor APIs (Native backend), but it can also query SDKMAN.

```{r}
# List versions from native backends (Corretto, Temurin, Zulu)
java_list_available()

# List versions from SDKMAN
java_list_available(backend = "sdkman")
```

# 3. Download `Java` JDK distributions

To download a specific Java distribution, use the `java_download` function:


```{r}
java_17_distr <- java_download(version = 17)
java_17_distr
```

Expected output (on a Windows machine):

```
Detected platform: windows
Detected architecture: x64
You can change the platform and architecture by specifying the `platform` and `arch` arguments.
Downloading Java 17 (Corretto) for windows x64 to C:\Users\user_name\AppData\Local/R/cache/R/rJavaEnv/distrib/amazon-corretto-17-x64-windows-jdk.zip
 [100%] Downloaded 187323560 bytes...
Download completed.
MD5 checksum verified.
[1] "C:\\Users\\user_name\\AppData\\Local/R/cache/R/rJavaEnv/distrib/amazon-corretto-17-x64-windows-jdk.zip"
```

This function will download the specified version of `Java` JDK (currently defaults to the only supported Corretto distribution by Amazon, more will be added in new package releases).

The function returns the path to the downloaded distribution file, which is just a `zip` or `gz` archive.

By default, if no `cache_path` is specified when calling `java_download()`, the distribution is downloaded into the folder defined by `getOption("rJavaEnv.cache_path")`. So with this argument you can override the globally set cache folder location for this single download.

`java_download()` also allows you to define the operating system and processor architecture with function arguments `platform` and `arch` (please see the function documentation using `?rJavaEnv::java_download`). For example, this may be useful if you are packaging a Java distribution for Linux while using a Windows or macOS machine. This may also be useful if for some reason automatic OS and CPU detection fails.

For example, this line above will download the `Amazon Corretto` `Java JDK` version 8 for x86-64 Linux, even though you may be running this code on aarch64 macOS or x86_64 Windows machine:

```{r}
java_8_linux64_distr <- java_download(
  version = 8,
  platform = "linux",
  arch = "x64"
)
```

Example expected output (on a Windows machine):

```
Detected platform: linux
Detected architecture: x64
You can change the platform and architecture by specifying the `platform` and `arch` arguments.
Downloading Java 8 (Corretto) for linux x64 to C:\Users\user_name\AppData\Local/R/cache/R/rJavaEnv/distrib/amazon-corretto-8-x64-linux-jdk.tar.gz
 [100%] Downloaded 118096667 bytes...
Download completed.
MD5 checksum verified.
```


MD5 checksum verified.
```

## Using SDKMAN Backend

You can also download distributions using the SDKMAN backend. Note that SDKMAN does not provide checksums for verification.

```{r}
java_21_sdkman <- java_download(
  version = 21,
  distribution = "Temurin",
  backend = "sdkman"
)
```

Example expected output:

```
Detected platform: windows
Detected architecture: x64
Downloading Temurin Java 21...
 [100%] Downloaded 201696048 bytes...
! Skipping checksum (unavailable for SDKMAN)
```


# 4. Manage downloaded `Java` JDK distributions

You can manage downloaded distributions with the `java_list` and `java_clear` functions.


```{r}
java_list_distrib()
```

Expected output (on a Windows machine):

```
java_distr_path
1 C:\\Users\\user_name\\AppData\\Local/R/cache/R/rJavaEnv/distrib/amazon-corretto-17-x64-windows-jdk.zip
2 C:\\Users\\user_name\\AppData\\Local/R/cache/R/rJavaEnv/distrib/amazon-corretto-21-x64-windows-jdk.zip
3 C:\\Users\\user_name\\AppData\\Local/R/cache/R/rJavaEnv/distrib/amazon-corretto-8-x64-linux-jdk.tar.gz
```

```{r}
java_clear_distrib()
```

Example expected output (on a Windows machine):

```
i Existing Java distributions:
1: C:\Users\user_name\AppData\Local/R/cache/R/rJavaEnv/distrib/amazon-corretto-17-x64-windows-jdk.zip
2: C:\Users\user_name\AppData\Local/R/cache/R/rJavaEnv/distrib/amazon-corretto-21-x64-windows-jdk.zip
3: C:\Users\user_name\AppData\Local/R/cache/R/rJavaEnv/distrib/amazon-corretto-8-x64-linux-jdk.tar.gz
i Enter the number of the distribution to delete, 'all' to delete all, or '0' or any other character to cancel:
```

You can also delete all downloaded distributions without consent with:

```{r}
java_clear_distrib(delete_all = TRUE)
```

Example expected output:

```
Java distributions cache cleared.
```

# 5. Install from downloaded `Java` JDK distributions into current project

As we have cleared all downloaded `Java` distributions, let us re-download a few of them:

```{r}
java_8_distr <- java_download(8)
java_17_distr <- java_download(17)
java_22_distr <- java_download(22)
```


Now that we have downloaded the `Java` distributions, we can install any one of them into the project. Installation in the context of `rJavaEnv` means extracting the distribution into the `installed` cache folder (which is in the same root cache directory as the zip archives of the distributions), and linking the files (via symlinks in `macOS` and `Linux`, and junctions in `Windows`[^1]). If we set the `autoset_java_env` argument to `FALSE` we will need to set the `JAVA_HOME` and `PATH` environment variables manually.

[^1]: In Windows, it is sometimes impossible to create junctions for files and folders. In this case, the `Java` installation will have to be copied into the project folder, which will take some time. This behaviour is the same `renv`'s for R packages and is an operating system limitation.


```{r}
java_install(
  java_distrib_path = java_17_distr,
  autoset_java_env = FALSE
)
```

Example expected output (on a Windows machine):

```
Java 17 (amazon-corretto-17-x64-windows-jdk.zip) for windows x64 installed at C:\Users\user_name\AppData\Local/R/cache/R/rJavaEnv/installed/windows/x64/Corretto/native/17 and symlinked to
C:/Users/user_name/AppData/Local/Temp/75/Rtmp0MasTW/rjavaenv/windows/x64/Corretto/native/17
```

If we did leave `autoset_java_env = TRUE` which is a default option, `install_java()` function would automatically set the `JAVA_HOME` and `PATH` environment variables manually in both the current `R` session and in the current project/working directory by writing a few lines into the `.Rprofile` file in the current working directory. This way each time you would start an R session, the `JAVA_HOME` and `PATH` environment variables would be set to the correct path to the installed `Java` distribution.

Alternatively, you may want to not install `Java` into your project, but just unpack the distribution into the cache folder. For this, you can use the `java_unpack()` function, passing it the path to the downloaded `Java` distribution from before:

```{r}
java_home <- java_unpack(java_distr_path = java_8_distr)
```

`java_unpack` will return full path to the unpacked `Java` distribution that you can use to set the environment variables in the section below using `java_env_set()`.


# 6. Manually set installed `Java` environment

Let us install another downloaded `Java` distribution, this time with `autoset_java_env = TRUE`:

```{r}
java_install(
  java_distrib_path = java_8_distr,
  autoset_java_env = TRUE
)
```

Expected output (on a Windows machine):

```
Java distribution amazon-corretto-8-x64-windows-jdk.zip already unpacked at C:\Users\user_name\AppData\Local/R/cache/R/rJavaEnv/installed/windows/x64/Corretto/native/8
v Current R Session: JAVA_HOME and PATH set to C:\Users\user_name\AppData\Local/R/cache/R/rJavaEnv/installed/windows/x64/Corretto/native/8
v Current R Project/Working Directory: JAVA_HOME and PATH set to 'C:\Users\user_name\AppData\Local/R/cache/R/rJavaEnv/installed/windows/x64/Corretto/native/8' in .Rprofile at 'C:/Users/user_name/AppData/Local/Temp/75/R
tmp0MasTW'
Java 8 (amazon-corretto-8-x64-windows-jdk.zip) for windows x64 installed at C:\Users\user_name\AppData\Local/R/cache/R/rJavaEnv/installed/windows/x64/Corretto/native/8 and symlinked to
C:/Users/user_name/AppData/Local/Temp/75/Rtmp0MasTW/rjavaenv/windows/x64/Corretto/native/8
```

We can now list the `Java` versions installed in the project directory:

```{r}
java_list_project()
```

Expected output (on a Windows machine):

```
path platform arch distribution backend version
1 C:/Users/user_name/AppData/Local/Temp/75/Rtmp0MasTW/rjavaenv/windows/x64/Corretto/native/17  windows  x64    Corretto  native      17
2  C:/Users/user_name/AppData/Local/Temp/75/Rtmp0MasTW/rjavaenv/windows/x64/Corretto/native/8  windows  x64    Corretto  native       8
```

We first installed Java 17 with `autoset_java_env = FALSE`, and then installed Java 8 with `autoset_java_env = TRUE`. If we check which version is currently set as the `JAVA_HOME` environment variable, we would get the second installed `Java` version, which is Java 8.

```{r}
java_check_version_cmd()
```

Expected output (on a Windows machine):

```
JAVA_HOME: C:\Users\user_name\AppData\Local/R/cache/R/rJavaEnv/installed/windows/x64/Corretto/native/8
Java path: /c/Users/user_name/AppData/Local/R/cache/R/rJavaEnv/installed/windows/x64/Corretto/native/8/bin/java
Java version: "openjdk version \"1.8.0_422\" OpenJDK Runtime Environment Corretto-8.422.05.1 (build 1.8.0_422-b05) OpenJDK 64-Bit Server VM Corretto-8.422.05.1 (build 25.422-b05, mixed mode)"
[1] TRUE
```

To switch to Java 17 we need to manually set the `JAVA_HOME` and `PATH` environment variables with `java_env_set()`. To identify which path to pass to `java_env_set()`, we can use the `java_list("project")` function again:

```{r}
java_list_project()
```

Expected output (on a Windows machine):
```
path platform arch distribution backend version
1 C:/Users/user_name/AppData/Local/Temp/75/Rtmp0MasTW/rjavaenv/windows/x64/Corretto/native/17  windows  x64    Corretto  native      17
2  C:/Users/user_name/AppData/Local/Temp/75/Rtmp0MasTW/rjavaenv/windows/x64/Corretto/native/8  windows  x64    Corretto  native       8
```

As we can see, the Java 17 is currently number 1 in this list, so we save this path:


```{r}
java_home_17 <- java_list_project(output = "vector")[1]
java_home_17
```

Expected output (on a Windows machine):

```
"C:/Users/user_name/AppData/Local/Temp/75/Rtmp0MasTW/rjavaenv/windows/x64/Corretto/native/17"
```

Now we can reuse this path in the `java_env_set()` function:

```{r}
java_env_set(where = "both", java_home = java_home_17)
```

Expected output (on a Windows machine):

```
v Current R Session: JAVA_HOME and PATH set to C:/Users/user_name/AppData/Local/Temp/75/Rtmp0MasTW/rjavaenv/windows/x64/Corretto/native/17
v Current R Project/Working Directory: JAVA_HOME and PATH set to 'C:/Users/user_name/AppData/Local/Temp/75/Rtmp0MasTW/rjavaenv/windows/x64/Corretto/native/17' in .Rprofile at 'C:/Users/user_name/AppData/Local/Temp/75/R
tmp0MasTW'
```

And finally we can double check which version of `Java` is going to be picked up by the `Java`/`rJava`-dependent `R` package that you are going to use with:

```{r}
java_check_version_cmd() # for pacakges that use Java via commandline, like opentripplanner
java_check_version_rjava() # for packages that use Java via `rJava`, like r5r
```


Expected output (on a Windows machine):

```
JAVA_HOME: C:/Users/user_name/AppData/Local/Temp/75/Rtmp0MasTW/rjavaenv/windows/x64/Corretto/native/17
Java path: /c/Users/user_name/AppData/Local/Temp/75/Rtmp0MasTW/rjavaenv/windows/x64/Corretto/native/17/bin/java
Java version: "openjdk version \"17.0.12\" 2024-07-16 LTS OpenJDK Runtime Environment Corretto-17.0.12.7.1 (build 17.0.12+7-LTS) OpenJDK 64-Bit Server VM Corretto-17.0.12.7.1 (build
17.0.12+7-LTS, mixed mode, sharing)"
[1] TRUE

Using current session's JAVA_HOME: C:/Users/user_name/AppData/Local/Temp/75/Rtmp0MasTW/rjavaenv/windows/x64/Corretto/native/17
With the user-specified JAVA_HOME rJava and other rJava/Java-based packages will use Java version: "17.0.12"
[1] TRUE
```

# 7. Cleanup

If you do not want to use `rJavaEnv` anymore, please clear the cache folders before removing the package:

```{r}
java_clear_project(delete_all = TRUE)
java_clear_installed(delete_all = TRUE)
java_clear_distrib(delete_all = TRUE)
```

Also, clear the `.Rprofile` file in the projects there you used the package:

```{r}
java_env_unset()
```
