---
title: "Get started with rJavaEnv"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with rJavaEnv}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Quick Start with rJavaEnv

This vignette provides a quick start guide for using the `rJavaEnv` package.

## Installation

You can install the development version of `rJavaEnv` from GitHub:

```{r eval=FALSE}
devtools::install_github("e-kotov/rJavaEnv")
```

## Example Workflow

### 1. Quick Install Java

This will download and install Java 21 (default) in the current project directory, and set the Java environment:

```{r eval=FALSE}
library(rJavaEnv)
java_quick_install(21)
```

### 2. Step-by-Step Installation

If you prefer a more controlled installation process, you can follow these steps:

#### Download Java

Download Java 21 distribution (default):

```{r eval=FALSE}
java_distr_path_21 <- java_download(version = 21)
```

#### Install Java

Install the downloaded Java distribution:

```{r eval=FALSE}
java_home_path_21 <- java_install(java_distr_path_21)
```

### 3. Set Java Environment

Set the JAVA_HOME and PATH environment variables to the installed Java directory:

```{r eval=FALSE}
java_env_set(java_home_path_21)
```

### 4. Check Java Version

Check the installed Java version using terminal commands:

```{r eval=FALSE}
java_check_version_cmd()
```

Check the Java version using `rJava` in a separate R session:

```{r eval=FALSE}
java_version_check_rjava("/path/to/installed/java")
```
